---
title: "Validate PSMs based on spectral and identification features"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Validate PSMs}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{PSMatch}
    %\VignetteDepends{BiocStyle,Spectra,PTMods}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("PSMatch")`<br />
**Authors**: `r packageDescription("PSMatch")[["Author"]] `<br />
**Last modified:** `r file.info("PSM.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r setup, message = FALSE, echo = FALSE}
library("PSMatch")
library("Spectra")
```

# Installation instructions

To install the `PSMatch` package from Bioconductor, make sure you have the
`BiocManager` package, available from CRAN, and then run

```r
BiocManager::install("PSMatch")
```

Additionally, this vignette makes use of another Bioconductor package called
`Spectra`. You can install the `Spectra` package in the same manner:

```{r}
BiocManager::install("Spectra")
```

# Introduction

This vignette is one among several illustrating how to use the `PSMatch`
package, in conjuction with the `Spectra` package, focusing on validation
metrics that can be applied on instances of merged `Spectra` and `PSM` objects.
For a general overview of the packages, see the `PSMatch` and `Spectra` package
manual pages (`?PSMatch` and `?Spectra`) and references therein.

# Data

The data that is used here can be downloaded from the
[PRIDE project PXD039419](https://proteomecentral.proteomexchange.org/cgi/GetDataset?ID=PXD039419).

## Load spectra and identification data

Let's load the *mzML* data and identification data and merge them together into
one `Spectra` object. Make sure to include both the MS1 and MS2 spectra without
filtering for PSM ranks, FDR and/or decoy/targets as these are used by the
various validation metrics.

```{r load, message = FALSE}
sp <- Spectra("path/to/data/examples.mzML")
psms <- read.delim("path/to/data/examples.tsv")
## no real need to create PSM objects ? Apart from the usefulness of pre-named
## columns (peptide, fdr, etc) but how to include those in the Spectra object
## with these new names ?

head(sp$pkey <- paste0(basename(sp$dataOrigin),
                       sub("ˆ.+scan=", "::", sp$spectrumId)))
head(psms$pkey <- paste0(basename(psms$filename),
                         sub("ˆ.+scan=", "::", psms$scannr)))

sp <- joinSpectraData(
    sp, psms,
    by.x = "pkey")

sp$sequence <- sp$peptide
```

# Metrics

A brief explanation on what the metrics are can be found in the help page
(`?identifieR::validatePSM`) but we'll go over them with more detail in this
section.

## checkABpresence

The `checkABpresence()` function checks whether both a2 and b2 ions are matched
in the observed spectrum based on the identified peptide sequence. This metric
confirms the correct fragmentation of the precursor ion. Its absence however
should not inform us of a bad fragmentation per se.

The result is a boolean of length `x`, `x` being the `Spectra` object's length.
If `TRUE`, the a2-b2 couple is matched, if `FALSE` it isn't.

## checkXYpresence

This metric is similar to `checkABpresence()` but specified for the x2-y2 couple.

## checkOverlap

The `checkOverlap()` function is an indicator for the possible presence or
absence of a modification on the identified sequence. Suppose `"PEPTIDEK"` is
the identified sequence and only the fragments b2, b3, y2, y3, y4 are matched.

```{r overlap, echo = FALSE}
sp_overlap <- DataFrame(msLevel = 2L, rtime = 2345, sequence = "PEPTIDEK")
sp_overlap$mz <- list(c(75.048614501953, 81.069030761719, 86.085876464844,
                88.039, 158.089569091797, 163.114898681641,
                173.128, 177.098587036133, 227.10263,
                232.137542724609, 233.140335083008, 259.938415527344,
                260.084167480469, 276.15539, 282.680786132812,
                284.079437255859, 291.208282470703, 315.422576904297,
                317.22509765625, 324.15539, 362.211944580078,
                391.18233, 433.255004882812, 504.26639,
                549.305236816406, 593.217041015625, 594.595092773438,
                609.848327636719, 631.819702148438, 632.324035644531,
                632.804931640625, 640.8193359375, 641.309936523438,
                641.82568359375, 678.357238769531, 685.28007,
                733.28007, 782.33283, 851.384033203125,
                880.414001464844, 881.40185546875, 906.396433105469,
                938.445861816406, 1022.56658935547, 1050.50415039062,
                1059.82800292969, 1107.52734375, 1138.521484375,
                1147.51538085938, 1226.056640625))
sp_overlap$intensity <- list(c(83143.03, 65473.8, 192735.53, 3649178.5,
                       379537.81, 89117.58, 922802.69, 61190.44,
                       2984798.75, 294798.75, 111935.03, 402512.57,
                       117443.59, 890773.67, 391008.15, 55350.43,
                       209952.97, 37001.18, 439515.53, 1390584.47,
                       46842.71, 1015457.44, 419382.31, 890378.77,
                       444406.66, 58426.91, 46007.71, 58711.72,
                       80675.59, 312799.97, 134451.72, 151969.72,
                       1961975, 69405.76, 395735.62, 71002.98,
                       3215457.75, 136619.47, 166158.69, 682329.75,
                       239964.69, 242025.44, 1338597.62, 50118.02,
                       1708093.12, 43119.03, 97048.02, 2668231.75,
                       83310.2, 40705.72))
sp_overlap <- Spectra(sp_overlap)
plotSpectraPTM(sp_overlap, USI = FALSE, deltaMz = FALSE, main = "checkOverlap checks the overlap between matched b- and y ions")
```

There is no overlap between the b- and y ions highlighting this gap between b3
and y4. The fact that  no more y ions or b ions are matched could be due to a
modification starting from b4 and y5 (for instance phosphorylation of Threonine).
We would suggest to try one or multiple modifications and compare the spectrum
matching (with and without modification) to see if more fragments
can be matched. If none of these increase the fragment indices, it might be that
there is an unknown modification or that there are a lot of missing fragments,
coincidentally forming this gap.

```{r overlap_mod, echo = FALSE}
plotSpectraPTM(sp_overlap, USI = FALSE, deltaMz = FALSE, fixed_modifications = c(T = 79.966), main = "There is an overlap between fragments when a modification is applied")
```

On the contrary, if the identified sequence includes a modification (suppose
`"PEPT[+79.966]IDEK"`) yet there is no overlap between the b- and y-ions,
generating this typical gap, it could be that the modification is absent or that
another modification is actually present.

The function returns a boolean of length `x`, `x` being the `Spectra` object's length.
Set to `TRUE` is there is the overlap, further confirming the identification and
`FALSE` if there is no overlap: setting some ambiguity in the confidence of the
identification.

## checkXrea

This metric is defined by
[Na, Seungjin, and Eunok Paek. 2006](https://pubs.acs.org/doi/full/10.1021/pr0603248).
as the area between a diagonal and the cumulative intensity distribution divided
by the complete area under the diagonal. The idea is that Xrea measures how
concentrated the signal is in a spectrum. The more the signal is concentrated in
fewer peaks, the higher the Xrea. If all peaks share about the same intensity,
peaks do not distinguish from noise, the spectrum is of bad quality and Xrea = 0.

To cite the paper on why to avoid using relative intensities:

>Most existing methods of analysis use relative intensity,
defined as the raw intensity normalized by the intensity of the
most abundant peak or the total intensity of all the peaks in a
spectrum. But this relative intensity normalization has some
shortcomings. First, when the single most abundant peak is
very strong relative to the rest of the peaks, as observed quite
often, intensities of the remaining peaks may be hard to
distinguish from the background noise and can thus be ignored. Second, when raw
intensities are small over the entire spectrum, values of relative
intensities will be more or less the same, and the peaks may
therefore look indistinguishable

The function returns a numeric for which the higher the value, the better the
quality of the spectrum. Values range from 0 to 1 with 0 being that all peaks
have the same intensity.

## checkShiftConsistency

This metric is dependent on the presence of a PTM. The identified sequence can
have one or multiple modifications, suppose `"PEPT[+79.966]IDEK"` with a
modified Threonine at b4 or y5 and above. There are exactly 7 possible fragment
positions (b5, b6, b7, b8, y6, y7, y8) that would include the modification in
its fragment. The more of these fragments are matched, the more confidence we
have in the identification.

Were there 2 modifications in the sequence, the number of possible fragments with
a shift would increase. Take `"PEPT[+79.966]ID[-17.974]EK"` for instance.
The potential fragments with modifications are now: b5, b6, b7, b8, y3, y4, y5,
y6, y7, y8. Some of these fragments have two modifications, which would further
increase the confidence of the identification, but this is not yet implemented
in the function.

The function returns a numeric of length equal to the `Spectra` object length
and `NA` if there is no modification in the identification. The value represents
the percentage of potential modified fragments that were matched. Higher
percentages increase the confidence in the identification.

## checkParentIonIntensity

The parent ion is the fragment with the same mass-over-charge as the precursor
ion. The parent ion represents the portion of the precursor ion that was not
fragmented. This lack of fragmentation is an indicator of bad spectrum quality.
Bad fragmentation is thus checked for by checking the parent ion intensity.
Generally, a good fragmentation would result in a relative intensity of around
10% of the base peak's intensity. This value is not set in stone and could vary
depending on the data.

The function returns a numeric that reflects the intensity of the
parent ion over the intensity of the base peak. If this value reaches 100%, the
parent ion is also the most intense peak in the spectrum and highlights incomplete
fragmentation.

## checkPrecursorPurity

This metric is based off [this paper on precursor purity](https://pubs.acs.org/doi/10.1021/acs.analchem.6b04358)
and uses `msPurity::pcalc()` to calculate the purity of the precursor ion used
to generate the fragments. It depends on the width of the isolation window. It
returns a list of length n containing numeric vectors with two elements,
n being the `Spectra` object length. The first element of the numeric is a
percentage obtained by dividing precursor ion intensity with the sum of peak
intensities within the tolerance window. The second element depicts how many
peaks there were in said window with a relative intensity higher than 10%.

## validatePSM

`validatePSM()` applies all the metrics mentioned above on a `Spectra` object of
any length. It computes the necessary `peaksData` and `labelFragments`
only once to avoid repetitions of these time-consuming functions.

The function returns a `dataframe` with the expected values from the respective
metrics for each peptide. The spectrum identifier (`pkey` in our case) is also
added, as well as the identification and its canonical counterpart and the
corresponding FDR so to match these values back to the identification data or
`Spectra` object.

# Analyse metric results

How well do these metrics actually fare compared to the well-known FDR for each
PSM ? Are there conclusions that can be drawn ? Is there some correlation
between some of these metrics and the calculated FDR ? Do these metrics merely
help to confirm identifications rather than refute some ?

This section will answer these questions.

## Run simulation

Let's first apply the `validatePSM()` function on our data. This may take a
while... grab some coffee !

```{r val_data, eval = FALSE}
res <- validatePSM(sp)
```

```{r load_res, echo = FALSE}
load("/storage/research/dduv/cbio-lg/user/deflandreg/phd-guillaume-deflandre/psm-validation/validatePSMs_results.rda")
```

Remember, we can match the results with the original identifications from our
PSM object using the "pkey".

```{r merge}
results <- merge(data.frame(psms), res, by = "pkey")
```

## FDR - metric relationships

Let's see whether there is some correlation between the metrics and the
calculated FDR.

### Differentiate boolean from numeric metrics

```{r corr_plots}
results$isDecoy <- ifelse(results$label, "Decoy", "Target")

numeric_metrics <- c("xrea", "shift_consistency",
                     "parent_ion_int", "precursor_purity", "fragmented_peaks")

boolean_metrics <- c("by_overlap", "a2b2_presence", "x2y2_presence")

```

### Scatterplot of numeric metrics

#### All peptides

```{r scatt_all, echo = FALSE}
results %>%
  pivot_longer(all_of(numeric_metrics), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = value, y = fdr, color = isDecoy)) +
  geom_point(alpha = 0.2, size = 0.5) +
  facet_wrap(~ metric, scales = "free_x") +
  theme_bw()

results %>%
    pivot_longer(all_of(numeric_metrics), names_to = "metric", values_to = "value") %>%
    ggplot(aes(x = value, fill = isDecoy)) +
    geom_density(alpha = 0.4) +
    facet_wrap(~ metric, scales = "free") +
    theme_bw()
```

#### Only modified peptides

```{r scatt_mod, echo = FALSE}
results[results$peptide.y != results$canonical_seq,] %>%
  pivot_longer(all_of(numeric_metrics), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = value, y = fdr, color = isDecoy)) +
  geom_point(alpha = 0.2, size = 0.5) +
  facet_wrap(~ metric, scales = "free_x") +
  theme_bw()

results[results$peptide.y != results$canonical_seq,] %>%
    pivot_longer(all_of(numeric_metrics), names_to = "metric", values_to = "value") %>%
    ggplot(aes(x = value, fill = isDecoy)) +
    geom_density(alpha = 0.4) +
    facet_wrap(~ metric, scales = "free") +
    theme_bw()
```

#### Only non-modified peptides

```{r scatt_no_mod, echo = FALSE}
results[results$peptide.y == results$canonical_seq,] %>%
  pivot_longer(all_of(numeric_metrics), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = value, y = fdr, color = isDecoy)) +
  geom_point(alpha = 0.2, size = 0.5) +
  facet_wrap(~ metric, scales = "free_x") +
  theme_bw()

results[results$peptide.y == results$canonical_seq,] %>%
    pivot_longer(all_of(numeric_metrics), names_to = "metric", values_to = "value") %>%
    ggplot(aes(x = value, fill = isDecoy)) +
    geom_density(alpha = 0.4) +
    facet_wrap(~ metric, scales = "free") +
    theme_bw()
```

### Boxplot for boolean metrics

#### All peptides

```{r box_box, echo = FALSE}
results %>%
  pivot_longer(all_of(boolean_metrics), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = as.factor(value), y = fdr, fill = isDecoy)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  # geom_boxplot(width = 0.2, outlier.size = 0.3, alpha = 0.6) +
  facet_wrap(~ metric) +
  theme_bw()
```

#### Only modified peptides

```{r box_box_mod, echo = FALSE}
results[results$peptide.y != results$canonical_seq,] %>%
  pivot_longer(all_of(boolean_metrics), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = as.factor(value), y = fdr, fill = isDecoy)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  # geom_boxplot(width = 0.2, outlier.size = 0.3, alpha = 0.6) +
  facet_wrap(~ metric) +
  theme_bw()
```

#### Only non-modified peptides

```{r box_box_non, echo = FALSE}
results[results$peptide.y == results$canonical_seq,] %>%
  pivot_longer(all_of(boolean_metrics), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = as.factor(value), y = fdr, fill = isDecoy)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  # geom_boxplot(width = 0.2, outlier.size = 0.3, alpha = 0.6) +
  facet_wrap(~ metric) +
  theme_bw()
```

## 4-step methodology

Below we have applied the 4-step methodology demonstrated in Deflandre
Guillaume's master thesis. For more details on the used functions, please refer
to said thesis or use `?identifieR::featureAnalysis()`.

As checkABpresence, checkXYpresence and checkOverlap are not continuous, these
metrics are exempt from the feature analysis. For the shift consistency, only
those peptides of interest (those identified with a modification) were kept.

### checkXrea

Good weight, poorly correlated but not significant in linear regression.

```{r feat_xrea}
featureAnalysis(results, "xrea")
```

### checkShiftConsistency

Poor weight, poorly correlated and not significant in linear regression.

```{r feat_shift}
featureAnalysis(results[results$peptide.y != results$canonical_seq,], "shift_consistency")
```

### checkParentIonIntensity

Poor weight, poorly correlated yet significant in linear regression, although
relatively low estimate.

```{r feat_pion}
featureAnalysis(results, "parent_ion_int")
```

### checkPrecursorPurity

Light weight, lightly correlated and unsignificant in linear regression.

```{r feat_precursor_purity}
featureAnalysis(results, "precursor_purity")
```

# Session information

```{r si}
sessionInfo()
```